The cardinality-constrained college application problem: The greedy algorithm and its implications for student welfare

### 2022-07-15 paper_alma/body.tex
discuss regression studies on fee waiver interventions

We show that the optimal portfolios for the cardinality-constrained college are \emph{nested} in the budget constraint, meaning that when $\mathcal{X}_h$ is the optimal portfolio of size $h$, $\mathcal{X}_h \subset \mathcal{X}_{h+1}$. The nestedness property is equivalent to the optimality of the \emph{greedy} algorithm that iteratively adds to $\mathcal{X}$ the school that elicits the greatest increase in the objective value. The proof of the nestedness property yields insight into the properties of the optimal college application strategy: Students who cannot afford to apply to many schools allocate a greater proportion of their application budget to safety schools than students with a large application budget. A classic finding of behavioral psychology is that agents with limited resources tend to make more risk-averse decisions than their wealthier peers, especially with respect to education and career choice \cite{hartlaubandschneider2012,vanhuizenandalessie2019}. In the case of college application, our model (though theoretical) provides a rational basis for this tendency.

@@

\subsection{Literature review}

\subsubsection{Related optimization problems}

@@

\subsubsection{Admissions markets}

@@

Let us discuss a few intuitive heuristics for the college application problem, how their
accuracy compares to the optimal solution, and the implications for student welfare.

@@

\subsection{Implications for student welfare}

% Clearly, individual students benefit from reducing barriers to application, but what are the tradeoffs, and how can we calculate them?

@@

In the college application problem, the value $t_0$ of the noncollege option can function as a risk-aversion parameter. However, our analysis of the distributional heuristic reveals that regardless of their underlying risk preferences, students must modulate their risk allocation---that is, the relative share of reach and safety schools in their portfolio---according to the number of schools to which they can afford to apply. Students with a large application budget can tolerate, and indeed benefit, from taking larger risks than those without. This means that in college application, variation in students' risk allocation arises from a combination of variance in underlying preferences and variance in students' application budgets. This property, which we may call \emph{optimal risk elasticity,} is not present in the Markowitz model.

From a welfare standpoint, this optimal risk elasticity offers weak evidence for the notion that reducing the costs of college application could improve student equity. On the other hand, dissuading students from applying to too many reach schools may be a desirable feature in an admissions market inasmuch as it eliminates the administrative costs associated with reading low-quality applications.

A further complication arises in an intrinsic feature of submodular set functions such as the objective function of the college application problem---that they exhibit rapidly diminishing returns: The optimality of the greedy algorithm implies that the gain in student utility from increasing the budget from $h-1$ to $h$ is larger than that from increasing it from $h$ to $h+1$.

@@

The fact that the optimal objective value grows sublinearly in $h$ may be relevant in the design of admissions markets. In the South Korean college admissions process, students are allowed to apply to only $h = 3$ colleges in the main application cycle, a policy which is thought to promote fairness. Suppose that utility is measured in monetary units, $t_0 = \$100$, and $v(\mathcal{X}_3) = \$130$ for a typical student. Then since $v(\mathcal{X}_4) \leq \$140$, a policy that increments the application limit will be unpopular if its tax cost exceeds $\$10$.

This also suggests, paradoxically, that colleges should offer application fee waivers to students who have already applied to many competitor schools. % interesting tradeoff

% you should increase $h$ until the point where the maximum marginal value of an additional application, namely $v(h) / h$, is small enough that students won't be tempted to spam reach schools with extra applications.

>>>>

We show that the optimal portfolios for the cardinality-constrained college are \emph{nested} in the budget constraint, meaning that when $\mathcal{X}_h$ is the optimal portfolio of size $h$, $\mathcal{X}_h \subset \mathcal{X}_{h+1}$. The nestedness property is equivalent to the optimality of the \emph{greedy} algorithm that iteratively adds to $\mathcal{X}$ the school that elicits the greatest increase in the objective value. The proof of the nestedness property yields insight into the properties of the optimal college application strategy: Students who cannot afford to apply to many schools allocate a greater proportion of their application budget to safety schools than students with a large application budget.% rewrite: A classic finding of behavioral psychology is that agents with limited resources tend to make more risk-averse decisions than their wealthier peers, especially with respect to education and career choice \cite{hartlaubandschneider2012,vanhuizenandalessie2019}. In the case of college application, our model (though theoretical) provides a rational basis for this tendency.

@@

% \subsection{Literature review}

\subsection{Related optimization problems}

@@

\subsection{Admissions markets}

@@

In this section, we discuss a few intuitive heuristics for the college application problem and how their
accuracy compares to the optimal solution. The analysis of these heuristics provides an alternative explanation for findings from the education literature showing that low-income students use a more risk-averse college application strategy than their wealthy peers.

@@

\subsection{Risk management and student welfare}

Further analysis of the optimal application strategy provides a novel explanation of findings from regression studies of college applicant behavior and speculate about policy interventions to improve student equity and welfare. We conjecture that broad policies to decrease the costs of application may yield a greater increase in student equity than interventions that narrowly target high-achieving, low-income students.

@@

In the college application problem, the value $t_0$ of the noncollege option can function as a risk-aversion parameter. However, the analysis of the distributional heuristic reveals that regardless of their underlying risk preferences, students must modulate their risk allocation---that is, the relative share of reach and safety schools in their portfolio---according to the number of schools to which they can afford to apply. Students with a large application budget can tolerate, and indeed benefit, from taking larger risks than those without. This means that in college application, variation in students' risk allocation arises from a combination of variance in underlying preferences and variance in students' application budgets. This property, which we may call \emph{optimal risk elasticity,} is not present in the Markowitz model.

A classic finding of behavioral psychology is that agents with limited resources tend to make more risk-averse decisions than their wealthier peers, especially with respect to education and career choice \cite{hartlaubandschneider2012,vanhuizenandalessie2019}. More concretely, wealthy students tend to apply to more reach schools than low-income students of similar academic ability, a tendency that some education scholars attribute to a lack of teachers or older peers who attended selective universities \cite{hoxbyandavery2012}. But given that low-income students tend to apply to fewer schools overall, optimal risk elasticity provides a rational explanation for the difference between their risk allocation and that of their wealthy peers. Hence, the failure of low-income students to apply to reach schools is not necessarily suboptimal decisionmaking due to risk aversion, but could be \emph{optimal} decisionmaking under a smaller budget. If the latter explanation is correct, then optimal risk elasticity suggests that reducing the costs of college application for all students could improve student equity without the need to overcome psychological factors such as risk aversion and the absence of high-achieving role models.

Further inspiration for policy interventions to improve student welfare arises in an intrinsic feature of submodular set functions: That they exhibit rapidly diminishing returns. The optimality of the greedy algorithm implies that the gain in student utility from increasing the budget from $h-1$ to $h$ is larger than that from increasing it from $h$ to $h+1$.

@@

The fact that $v(\mathcal{X}_h)$ grows sublinearly in $h$ may provide an economic explanation for the underwhelming efficacy of nudges, such as application fee waivers, designed to encourage talented low-income students to apply to selective universities \cite{gurantzetal2021}. In our model, the marginal value of an additional college application is greatest to students who have both a small application portfolio, and small admissions odds at the schools they are currently applying to---hat is, students whose portolio allocation is skewed toward reach schools. However, fee-waiver interventions have typically targeted high-achieving, low-income students (so-called ``one-offs'' \cite{hoxbyandavery2012,hoxbyandturner2013}) who already have high admissions odds at the schools in their portfolio due a combination of underlying academic talent and the aforementioned tendency of low-income students to favor safety schools. Therefore, our analysis suggests that the students who stand to benefit the most from nudges such as fee waivers are not high-achieving, low-income students, but rather students of middling ability who are prepared for college but unlikely to be admitted to every school.

%On the other hand, dissuading students from applying to too many reach schools may be a desirable feature in an admissions market inasmuch as it eliminates the administrative costs associated with reading low-quality applications. Our analysis of student utility also can inform the design of centralized admissions markets. In the South Korean college admissions process, students are allowed to apply to only $h = 3$ colleges in the main application cycle, a policy which is thought to promote fairness. Suppose that utility is measured in monetary units, $t_0 = \$100$, and $v(\mathcal{X}_3) = \$130$ for a typical student. Then since $v(\mathcal{X}_4) \leq \$140$, a policy that increments the application limit will be unpopular if its tax cost exceeds $\$10$.
